A Shell-like Model for General Purpose Programming

### 2019-01-18 doc/paper.tex
paper updates

% notes
\documentclass[english,PRO]{ipsj}
\usepackage{PROpresentation}
\PROheadtitle{y-n-(x): Manuscript for presentation at IPSJ-SIGPRO, d m Y.}

@@

\usepackage{textcomp}
\usepackage{comment}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphics,graphicx}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{auto-pst-pdf}

@@

\begin{document}
\title{A Shell-like Model for General Purpose Programming}

\affiliate{TITech}{%東京工業大学情報理工学院数理・計算科学系 \\
Department of Mathematical and Computing Science, Tokyo Institute of Technology,
Meguro, Tokyo 152-8552}
\affiliate{KTH}{Kungliga Tekniska H\"ogskolan\\
KTH Royal Institute of Technology, Stockholm, Sweden, 10044}

\author{Jeanine Miller Adkisson}{TITech}[jneen@jneen.net]
\author{Johannes Westlund}{TITech,KTH}[jwestlun@kth.se]
\author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]

@@

\begin{abstract}
Shell languages such as bash are designed to integrate with an OS, which mainly involves managing processes with implicit input and output streams. They also attempt to do this in a compact way that could be reasonably typed on a command-line interface.

@@

By way of a new programming language, \emph{Magritte}, we propose a general-purpose programming language with semantics similar to bash. In this paper, we discuss the early design of such a system, in which the primary unit of composition, like bash, is processes with asynchronous inputs and outputs, which can be read from or written to at any time, and which can be chained together via a pipe operator. We also explore concurrency semantics for such a language.
\end{abstract}

\maketitle

@@

\noindent
The UNIX shell programming model has been shown to be a successful model for composing programs---spawning unrelated programs in parallel to communicate over operating-system pipes. Beyond the most simple tasks, however, \verb/bash/ and similar tools break down, due to the byte-based nature of the system: variables can only hold byte arrays, and pipes can only manage raw byte streams, with no boundaries between values.

In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes which communicate through synchronous channels or pipes in an ad-hoc manner --- and in which this facility is the primary method of composing different functions or units.

@@

\subsection*{Outline}\noindent
In this work, we present a new shell language \emph{Magritte}. Section 2 describes design requirements for a language in this space. Section 3 and 4 describe Magritte's design and implementation. Section 5 describes our future goals with this project, and in Section 6 we review various related works and alternative approaches.

@@

\subsection{Programming With Values}
\subsubsection{Not Everything Is a String}\noindent
The first core issue that prevents most shells from being viable for large programs is that only strings are allowed as first-class values.

Several projects\cite{haahr,duff}, including bash itself, have attempted to add lists and hash maps as standard objects. Unfortunately, while they allow these values to be stored in variables, most systems do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
Since values in these languages cannot be passed through pipes, they are relegated to a second class of value, one for which pipe-based composition is not possible.

Thus if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through channels.

\subsubsection{Consistent Channels}\noindent
Since UNIX pipes can only manage bytes, usually channels with a fixed buffer are used for performance reasons. However, it is important if a pipe is to be used for composition that reads correspond exactly to writes, and that data does not become corrupted by interleaving.

For example, consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and the values are collected in a single output. This might be expressed in bash as:
\noindent\begin{lstlisting}[morekeywords={do,done}]
# <@\underline{\color{gray} consume}@> the stream, labeling every line
label() { while read x; do echo "$1$x"; done ;}

# process the stream in the background
# with two threads
split() { label a & label b & ;}

# <@\underline{\color{gray} produce}@> and process the numbers 1-100,
# limiting output to the first 3 lines
seq 100 | split | head -3
\end{lstlisting}

\noindent
With this code, a user might expect the output to be three lines, each consisting of a letter \verb/a/ or \verb/b/, and a number, for example:

@@

\noindent
Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved, resulting in outputs such as:

\noindent\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
a2
b1
b
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
b12
a
a4
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
a
b12
a34
\end{lstlisting}
\end{minipage}

@@

\noindent
In order to provide a channel implementation that is usable for large programs, we must ensure that it is \emph{consistent}---that is, that every read corresponds to exactly one write.

\subsubsection{Return Semantics}\label{return-semantics}\noindent
Pipeline elements such as these do not generally return values---instead, they output data to their output stream. However, for serious programs it is desirable to capture some kind of return value from functions. Es Shell\cite{haahr} accomplishes this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.

If we allow values to be written to output streams, however, this approach is redundant---a function author must decide whether to return a value or write it to the standard output. Thus it is desirable to have an interpolation syntax which \emph{collects} the output of a command synchronously.

A priori, such a collection mechanism would have to return an list, as any function call may output zero or more values. However, that approach would require the callers of functions to manually unwrap lists on every function call.

\subsection{Language Features}
\subsubsection{Variable Scope}\noindent
Like the Es shell, we also wish to provide lambda functions with lexically scoped variables. However, Es has an ambiguity with respect to its variables---since both lexically scoped variables and dynamically scoped variables share the same syntax, the scope of a variable depends on which variables happen to be bound in the function's definition environment. Therefore a simple mistype of a variable name will result not in an error, but in the variable silently becoming dynamically bound.

\subsubsection{Objects}\noindent
For large programs, an object system is desirable as a way to provide user-defined types and abstractions. The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer. Therefore, if we allow environments as first-class values, they can be used in a straightforward manner as prototype-based objects.

\subsubsection{Trees and Pattern Matching}\noindent
For programs that deal with syntax or other kinds of trees, it is convenient to have a compact tree representation, which can be made possible with nestable vectors. It is also desirable to have some mechanism for pattern matching against these kinds of values. These kinds of structures are impossible using the list features in systems such as RC Shell\cite{duff} and Es Shell, which implicitly flatten all lists.

\subsection{Automatic Process Cleanup}
\subsubsection{Interruption}\noindent
A core requirement of pipe-based languages is a semantic of process \emph{interruption}: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
\begin{lstlisting}
  read-lines tmp/large-file                        <@\color{gray}\hfill $(1)$@>
    | take 10                                      <@\color{gray}\hfill $(2)$@>
    | each (?line => do-expensive-work <@\$@>line) <@\color{gray}\hfill $(3)$@>
\end{lstlisting}

\noindent
In this example, three processes are spawned: (1) a process with an open file that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.

In a pipe-based language, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
A user will also expect that, after the first 10 lines are processed and the \verb/take/ function returns, the file will be closed, and all three processes exit.
This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.

In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
Similarly, the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.

\subsubsection{Compensation}\noindent
In order to observe interruption, user code must be able to react to and handle interruption. The most direct way to observe interruption is through a system of \emph{compensation actions}, which are expressions which will be evaluated in the case of an interruption.

Compensations are the only way in which interruptions can be directly observed by user code---without compensations, the only way of observing forever-blocked processes would be to inspect the process table, or to observe the memory footprint of the program. Thus, we can define the primary task of the interruption system as \emph{running a process's compensation actions at the appropriate time}.

\subsubsection{Lazy Interruption vs. Eager Interruption}\label{lazy-interruption}\noindent
In designing an interruption semantic, we must first decide whether processes are eagerly interrupted when their channels close, or only interrupted on channel operations.

@@

\noindent
The first process will write three times, and then continue to do other processing in-thread that does not write to the standard output.
The final \verb/take 3/ operation will return after 3 inputs are read.
It is important to decide, then, whether the first process should be \emph{interrupted} in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or whether it should be left alone until it attempts to write a value (\emph{lazy interruption}).

An advantage of lazy interruption is that a user can identify precisely which points in the code have the potential to be interrupted---those points which run a \verb/put/ or a \verb/get/. Whereas with eager-interruption semantics any point in the code may be interrupted, introducing the need for users to either mark critical uninterruptible sections or be very careful about setting correct compensations in the case of mutative algorithms.

On the other hand, lazy interruption has the disadvantage that a producing process must do enough work to produce one more value than will be consumed. If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in doing a large amount of unnecessary work.

\subsubsection{Non-closing of Channels}\label{non-closing}\noindent
Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel. It would not be appropriate in this case to interrupt other processes attached to the channel, as they are still able to communicate.

We define our guiding principle for the appropriate time to interrupt a process as: \emph{A process is interrupted exactly when it can no longer be woken up}.
When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
Therefore the appropriate time for it to be interrupted is when it is blocked on a channel that will never receive any more operations.
In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.

Luckily, in pipe-based languages, the arrangement of pipes is usually specified at process spawn time. This enables us to maintain a process register inside of each channel, so that we can decide when all readers or all writers have detached from a channel, at which time the channel closes and interrupts all blocked processes.

\subsubsection{Closing Permanence}\label{closing-permanence}\noindent
Even when we close a channel when all processes on the read or write sides close, there is always the possibility of new processes spawning and attempting to connect to a channel. In such a case, some systems leave open the possibility of \emph{reopening} a channel for further use. In UNIX, for example, the named-pipes API allows for persistent channels \cite{linux-pipe} that, being merely files, can be reopened at any time.

However, this can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:
\begin{lstlisting}
c = (make-channel)
& count-forever > $c
& take 10 < $c
put 10 > $c
\end{lstlisting}

\noindent
This example, while contrived, represents an unavoidable race with first-class channels---whether the process running \verb/take 10/ closes the channel before or
after the \verb/put 10/ invocation opens the channel for writing. If we allow channel reopening, we will either block forever, or insert the number 10 into the stream, depending on which happens first. However if we do not allow channel reopening, we can say that, if a process initiates a read or write on a closed channel, it is immediately interrupted. In this case, both sides of the former race have the same termination behavior---the process is closed when \verb/take 10/ returns.

\subsubsection{Masking Interruptions}\label{masking}\noindent
A user will often wish to continue after a particular output channel has closed, and in this case will need a facility to mask exceptions for particular channels. Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
\begin{lstlisting}
(sample) = (
  hold = (get)
  i = 1
  each (?v =>
    $i = (inc $i)
    # mutate the $hold variable
    # with probability 1/$i
    prob (reciprocal $i) 1 && ($hold = %v)
  )

  put $hold
)
\end{lstlisting}

\noindent
In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability. After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.

However, with the semantics described above, the \verb/each/ function will loop until it receives an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line. In order to avoid this situation, the \verb/each/ function must somehow indicate that it intends to consume the entire stream, and handle the interrupt signal.

@@

\subsubsection{Lexical vs. Dynamic Variables}\noindent
In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/). Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g. \verb/x = 1/), so that we can maintain compatibility with standard environment files. Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g. \verb/?x/) to allow for a planned extension for pattern matching.

@@

\subsubsection{Lambda Functions}\noindent
Lambda functions are specified with parenthesized expressions containing the arrow symbol \verb/=>/, which separates the bindings from the body. Multiple matching clauses are possible, and will be extended in the future with more general pattern matching. Clauses are separated at the beginning of lines that contain \verb/=>/, which avoids the need for any indentation-based parsing. For example:
\begin{lstlisting}
my-function = (
  ?x =>
    put one-argument %x
  ?y ?z =>
    put two-arguments
    put %y %z
  ... => ...
)
\end{lstlisting}

@@

\noindent
Functions are scanned statically for free lexical variables, which are captured at function definition time, and merged into the environment at call time.

\subsubsection{Nestable Vectors}\noindent
We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/. Combined with bareword strings, we can represent trees in a straightforward manner:
\begin{lstlisting}
a-tree = [node [node [leaf 1] [leaf 2]] [leaf 3]]
\end{lstlisting}
\noindent
We discuss a pattern matching extension in \emph{Future Work}. A typical strategy for traversing this kind of structure might be a function that put all leaf node values to its output in a predefined order, to be consumed by another process.

The builtin function \verb/for/ takes a vector as an argument and outputs each element in order, so that a normal vector can be traversed by the pattern:
\begin{lstlisting}
for [1 2 3 4 5] | each (?el => ...)
\end{lstlisting}

\subsubsection{Environments as Objects}\noindent
We support environment capture with the straightforward \verb/{ }/ syntax. The semantics of curly braces are to run the code as normal, then detach the running environment from its parent and return it. Therefore all assignment syntaxes are available, including function definition. For example:
\begin{lstlisting}
(make-account ?balance) = {
  balance = $balance
  (deposit ?amt) =
    (%balance = (add %amt %balance))
  (withdraw ?amt) = (
    (%balance = (sub %amt %balance))
}
\end{lstlisting}

\noindent
Environment lookup uses the \verb/!/ symbol, as in \verb/$env!key/. We use this symbol because it is already reserved by bash, unlike both \verb\/\ or \verb/./ which need to be available in bareword syntax to integrate with external programs. Environments can also be mutated using the same access syntax. For example:
\begin{lstlisting}
account = (make-account 10)
$account!withdraw 4
put $account!balance # => 6
\end{lstlisting}

\subsubsection{Reading and Writing}\noindent
Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively. These block until they complete a communication.

\subsubsection{Collection and Substitution}\noindent
In order to use output values seamlessly as return values without unwrapping arrays at every function call as described in Section \ref{return-semantics}, Magritte uses normal parentheses to collect and \emph{expand} the resulting values into the current command vector---increasing the argument number by the number of values output from the function. For example:
\begin{lstlisting}
# A function defintion: output three values
(count-three) = (put 1; put 2; put 3)

# Parentheses collect the three writes and expand
# 1 2 3 in-place
other-fn 0 (count-three) 4

@@

\noindent
These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that because we write transactional values to the output, we have the ability to properly separate values without relying on whitespace.

@@

\section{Future Work}\label{future-work}

>>>>

% default is sigpro, define \sigplan to switch

\ifx\sigplan\undefined
\newcommand{\ifsigplan}[1]{}
\newcommand{\ifsigpro}[1]{#1}
\else
\newcommand{\ifsigplan}[1]{#1}
\newcommand{\ifsigpro}[1]{}
\fi

\ifsigpro{ \documentclass[english,PRO]{ipsj} }
\ifsigplan{ \documentclass[sigplan,9pt]{acmart} }

@@

\usepackage{textcomp}
\usepackage{comment}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphics,graphicx}
\usepackage{pstricks,pst-node,pst-tree}
\usepackage{auto-pst-pdf}
% \usepackage[symbol]{footmisc}

@@

\ifsigpro{
\usepackage{PROpresentation}
\PROheadtitle{2018-4-(5): Manuscript for presentation at IPSJ-SIGPRO, 17 Jan 2019.}
}

\ifsigplan{
\acmDOI{}
\copyrightyear{2019}
\acmISBN{}
\acmConference[PASS Workshop]{}{April 2019}{Genova, Italy}
\setcopyright{none}
\titlenote{This work was presented at the 122nd IPSJ SIGPRO workshop on January 17, 2019. The presentation was not refereed, and the same manuscript was distributed only to the participants to the workshop.}
}

\begin{document}

\title{A Shell-like Model for General Purpose Programming}

\ifsigpro{
\affiliate{TITech}{%東京工業大学情報理工学院数理・計算科学系 \\
Department of Mathematical and Computing Science, Tokyo Institute of Technology,
Meguro, Tokyo 152-8552}
\affiliate{KTH}{Kungliga Tekniska H\"ogskolan\\
KTH Royal Institute of Technology, Stockholm, Sweden, 10044}

\author{Jeanine Miller Adkisson}{TITech}[jneen@jneen.net]
\author{Johannes Westlund}{TITech,KTH}[jwestlun@kth.se]
\author{Hidehiko Masuhara}{TITech}[masuhara@acm.org]
}

\ifsigplan{
\author{Jeanine Miller Adkisson}
\affiliation{%
  \institution{Tokyo Institute of Technology, Japan}
}
\email{jneen@jneen.net}

\author{Johannes Westlund}
\affiliation{%
  \institution{Tokyo Institute of Technology, Japan}
}
\affiliation{%
  \institution{KTH Royal Inst. of Technology, Sweden}
}
\email{jwestlun@kth.se}

\author{Hidehiko Masuhara}
\affiliation{
  \institution{Tokyo Institute of Technology, Japan}
}
\email{masuhara@acm.org}
}

@@

\begin{abstract}
Shell scripting languages such as bash are designed to integrate with an OS, which mainly involves managing processes with implicit input and output streams. They also attempt to do this in a compact way that could be reasonably typed on a command-line interface.

@@

By way of a new programming language, \emph{Magritte}, we propose a general-purpose programming language with semantics similar to bash. In this paper, we discuss the early design of such a system, in which the primary unit of composition, like bash, is processes with input and output channels, which can be read from or written to at any time, and which can be chained together via a pipe operator. We also explore concurrency semantics for such a language.
\end{abstract}

\settopmatter{printfolios=true}
\maketitle

@@

\noindent
The UNIX shell programming model has played an important role in integrating applications and operating systems by composing programs---spawning independent programs in parallel to communicate over operating-system pipes. Beyond the most simple tasks, however, bash and similar tools break down, due to various language deficiencies.

In this paper, we use the term \emph{pipe-based language} to mean a programming language intended to be used on a command line, with the ability to spawn many processes which communicate through synchronous channels or pipes in an ad-hoc manner---and in which this facility is the primary method of composing different functions or units.

@@

\subsection*{Outline}\noindent
In this work, we present a new pipe-based language \emph{Magritte}. Section 2 describes design requirements for a language in this space. Section 3 and 4 describe Magritte's design and implementation. Section 5 describes our future goals with this project, and in Section 6 we review various related work and alternative approaches.

@@

\subsection{Programming With Values}

\subsubsection{Value Pipes}\noindent
It is desirable in a general purpose language to be able to create and use complex data structures, and to use them freely throughout the language. In particular, if we are to allow for pipe-based composition to be the core composition method for large programs, we must allow rich values to be passed through pipes.

Several projects\cite{haahr,duff}, including bash itself, have attempted to add arrays and associative arrays as standard objects, but not as first class values: they do not allow these data structures to be passed into or returned from functions, or importantly, passed through pipes.
%These values are therefore relegated to a second class of value, one for which pipe-based composition is not possible.

Furthermore, serialization through pipes is generally not possible except in a single-threaded case, since complex data can be corrupted through interleaving.

For example, consider the fairly common architecture of producer/consumer: A \emph{producer} process produces values that are processed in parallel by multiple \emph{consumer} processes, and the values are collected in a single output. This might be expressed in bash as:
\begin{lstlisting}[morekeywords={do,done}]
# <@\underline{\color{gray} consume}@> the stream, labeling every line
label() { while read x; do echo "$1$x"; done ;}

# process the stream with two threads
split() { label a & label b & ;}

# <@\underline{\color{gray} produce}@> and process the numbers 1-100,
# limiting output to the first 3 lines
seq 100 | split | head -3
\end{lstlisting}
With this code, a user might expect the output to be three lines, each consisting of a letter \verb/a/ or \verb/b/, and a number, for example:

% \footnotetext{hello world}

@@

\noindent
Unfortunately, when we run this process, the outputs from the \verb/label/ function become interleaved\footnotemark, resulting in outputs such as:
\footnotetext{We have observed some behavior in the output that cannot be explained simply by interleaving, suggesting there may be some other race conditions in play.}

\noindent\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
a2
b1
b
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
b12
a
a4

\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.14\textwidth}
\begin{lstlisting}{Name}
a
b12
a34
\end{lstlisting}
\end{minipage}

@@

\noindent
In order to provide a channel implementation that is usable for large programs, we must ensure that it is \emph{consistent}---that is, that every read corresponds to exactly one write. Value pipes accomplish this by making a value the smallest atomic unit of communication.

% cat $(ls | grep ...)
\subsubsection{Capture and Substitution}\label{return-semantics}\noindent
For processes that output values, we need to support a mechanism to capture those values for use in variables, data structures, and function arguments, similar to backticks or \verb/$(...)/ in bash. This enables processes to be used like functions, which return data to their caller.

But since in most shells, rich values are not writable to output streams, the output cannot effectively be used as a return path for values. Es Shell\cite{haahr} accounts for this by introducing a \emph{return value} which is separate from stream output and can be accessed via special-purpose call syntax.

If we allow values to be written to output streams, however, we can directly capture values written to the output.

\subsubsection{Modern Language Features}
%Suggestion: without using subsubsections, list the features as a list (by using the itemize environment). And add first class functions into the list, and a prototype-based lookup mechanism.

Users will expect a modern programming language to have:
%Language features required for a modern general purpose language include:

\begin{itemize}
  \item \emph{Lambda functions with closure}.
    This requires the introduction of lexically scoped variables.
  \item \emph{Dynamic variables}.
     Most shell languages already include these, as OS Environment variables are dynamic by nature.
  \item \emph{Product structures with support for open recursion}.
    a prototype-based object system is sufficient for this.
  \item \emph{Sum structures}. In an untyped language, a method for pattern matching over nestable heterogenous lists is sufficient.
\end{itemize}

% For large programs, an object system is desirable as a way to provide user-defined types and abstractions. The existing foundations of shell languages already make use of the UNIX \emph{environments} API, which consists of a map of strings to strings, together with a parent pointer. Therefore, if we allow environments as first-class values, they can be used in a straightforward manner as prototype-based objects.

% For programs that deal with syntax or other kinds of trees, it is convenient to have compact nestable data structures that support pattern matching, which can be made possible with nestable vectors.

\subsection{Automatic Process Cleanup}
\subsubsection{Interruption}
\noindent
In shell programming, we tend to compose infinitely running processes together as pipeline elements. Thus we require a well-defined semantics of process \emph{interruption}\footnotemark: automatic clean-up of processes that will no longer be used. Consider the following Magritte code:
\footnotetext{We use the term \emph{interruption} in the generic sense---to mean the halting of normal flow in a process, due to some external event. It is unrelated to hardware or OS signalling.}
\begin{lstlisting}
  read-lines tmp/large-file                        <@\color{gray}\hfill $(1)$@>
    | take 10                                      <@\color{gray}\hfill $(2)$@>
    | each (?line => do-expensive-work <@\$@>line) <@\color{gray}\hfill $(3)$@>
\end{lstlisting}

\noindent
In this example, three processes are spawned concurrently: (1) a process with an open file that writes one line at a time to its output, (2) a process that reads 10 times from its input, writes each entry to the output, and then exits, and (3) a process that reads every input and calls a function to perform an expensive task.
A user's intent when typing such code may be to read 10 lines from a file and synchronously perform an action on each line.

A user will also expect that, after the first 10 lines are processed and the \verb/take/ function returns, the file will be closed, and all three processes exit.
This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is an infinite loop.

In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
Similarly, the call to the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.

\subsubsection{Compensation}\noindent
In interacting with an operating system, it is necessary to mange side effects, and gracefully recover or restore state in the case of an interruption. Such an error-handling system would also be a way for user code to directly observe interruption.

Without compensation of errors, the only way of observing forever-blocked processes would be to inspect the process table, or to observe the memory footprint of the program. Thus, we can define the primary task of the interruption system as \emph{running a process's compensation actions at the appropriate time}.

\subsubsection{Lazy Interruption vs. Eager Interruption}\label{lazy-interruption}\noindent
In a system such as this, there is a language design choice that must be made---whether interruption is \emph{eager}, in which processes are interrupted immediately upon closing a channel, or \emph{lazy}, in which processes are only interrupted upon interaction with a closed channel. Both approaches have advantages and drawbacks. In making this choice, we desire that the behavior of interruption be \emph{predictable}---however, there are two competing viewpoints for predictability.

@@

\noindent
The left-hand side of the pipe will write three times, and then continue to do other processing in-thread that does not write to the standard output.
The final \verb/take 3/ operation will return after 3 inputs are read.
It is important to decide, then, whether the process on the left should be interrupted in the middle of \verb/do-other-operation/ (\emph{eager interruption}), or whether it should be left alone until it attempts to write a value (\emph{lazy interruption}).

From the perspective of someone spawning a process, eager interruption can seem more predictable, as they can guarantee a point at which the process has stopped doing work.

However, from the perspective of a function author, lazy interruption is more predictable, because the author can identify precisely which points in the code have the potential to be interrupted---those points which run a \verb/put/ or a \verb/get/. Contrast this with eager-interruption semantics, where any point in the code may be interrupted, introducing the need for users to either mark critical sections and be very careful with implementing stateful algorithms.

On the other hand, lazy interruption has the disadvantage that a producing process must do enough work to produce one more value than will be consumed. If each value is relatively cheap to produce, this is not a problem, but in the case that the values are expensive to produce, this would result in a large amount of unnecessary work.

% sensei comments:
% I see. Let's delete this paragraph, and improve the first sentence.

% The problem was, while Section 2 is supposed to list the requirements, Section 2.3.3 merely points out existence of two viewpoints. It made me feel like "well I understand that you see two viewpoints, then which one you require?"

% I think the requirement wrt eager/lazy interruption is that the language should support both interruption semantics and let the programmer use one of them selectively.
\subsubsection{Closing of Channels}\label{closing}\noindent
Given that multiple processes may be reading from and writing to a channel, it is often the case that a communicating process will end when there are other processes still communicating over the channel. It would not be appropriate in this case to interrupt other processes attached to the channel, as they are still able to communicate.

We define our guiding principle for the appropriate time to interrupt a process as: \emph{A process is interrupted exactly when it can no longer be woken up}.
When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
Therefore the appropriate time for it to be interrupted is when it is blocked on a channel that will never receive any more operations.

This can be difficult to detect when a reference to a channel can be passed anywhere in the program as a standard value. Luckily, the arrangement of pipes and channels are usually specified at process spawn time. We can therefore relax our constraint to guarantee that channels will be closed at appropriate times in \emph{common architectures}, and that they never close if there are active readers or writers.

\subsubsection{Reopening Channels}\label{closing-permanence}\noindent
Some systems, like UNIX named pipes, allow a channel to be re-used by new processes after it has been closed\cite{linux-pipe}. This is, however, not a desirable feature, as it can lead to some unexpected races between a channel closing and a new process spawning. Consider the example:
\begin{lstlisting}
c = (make-channel)     # create a new channel
& count-forever > $c   # write infinitely
& take 10 < $c         # read 10 elements and exit
put 10 > $c            # write once from a new process
\end{lstlisting}

\noindent
This example represents an unavoidable race with first-class channels: between \verb/take 10/ closing the channel and \verb/put 10/ opening the channel for writing. If we allow channel reopening, we will either block forever, or insert the number 10 into the stream, depending on which happens first. However if we do not allow channel reopening, we can say that, if a process initiates a read or write on a closed channel, it is immediately interrupted. In this case, both sides of the former race have the same termination behavior---the process is closed when \verb/take 10/ returns.

\subsubsection{Masking Interruptions}\label{masking}\noindent
We must include a facility to control the extend of interruptions. This is because an interruption semantics can make it difficult to perform final calculations after a channel is closed.

% A user will often wish to perform final calculations after a particular output channel has closed, and interruption semantics can make this difficult.
Consider the following example, which sums all numbers from the standard input:
% Consider the following example, which uses reservoir-sampling to select a uniform random element from a stream:
% \begin{lstlisting}
% (sample) = (
%   hold = (get)
%   i = 1
%   each (?v =>
%     $i = (inc $i)
%     # mutate with probability 1/$i
%     prob (recip $i) 1 && ($hold = %v)
%   )
% 
%   put $hold
% )
% \end{lstlisting}

% \noindent
% In this example, the function \verb/each/ will consume the entire input stream and mutate a variable with a given probability. After the entire input stream is consumed, we wish to output the resulting value stored in the \verb/$hold/ variable.
\begin{lstlisting}
(sum) = (
  total = 0
  each (?x => %total = (add %total %x))
  put %total
)
\end{lstlisting}\noindent
In this example, the function \verb/each/ will consume the input stream and mutate a lexical variable (marked with \verb/%/). After the entire input stream is consumed, we wish to output the resulting \verb/%total/ value.

However, with the semantics described above, the \verb/each/ function will loop until it receives an interrupt signal from the input channel closing, which will interrupt the entire process and not continue to the following line.

@@

\subsubsection{Lexical vs. Dynamic Variables}\noindent
In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at variable reference and mutation points, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/). Let-binding is left unadorned (\verb/x = 1/), and will bind a variable both lexically and dynamically.

Let-binding uses this plain syntax to maintain compatibility with standard environment files. Variable binders in lambda arguments use \verb/?/ (e.g. \verb/?x/) to allow for unambiguous pattern matching, and are also bound both lexically and dynamically.

@@

% In order to support both lexical and dynamic variables, we introduce a separate syntax for lexically scoped variables at the variable reference point, using a \verb/%/ instead of \verb/$/ (for example, \verb/%x/). Let-binding will bind a variable both lexically and dynamically, and uses an unadorned variable name (e.g. \verb/x = 1/), so that we can maintain compatibility with standard environment files. Variable binders in lambda arguments are similarly both lexical and dynamic, and use \verb/?/ (e.g. \verb/?x/) to allow for unambiguous pattern matching.

% For mutation, we use assignment syntax, but with the location (dynamic or lexical) specified on the left-hand side, e.g. \verb/%x = 1/ or \verb/$x = 1/. In this way we avoid the need to declare variables, and allow ourselves to throw an exception when trying to mutate a non-existent variable, rather than silently creating a new local binding.
\subsubsection{Lambda Functions}\noindent
Lambda functions are specified with parenthesized expressions containing the arrow symbol \verb/=>/, which separates the bindings from the body. Multiple matching clauses are possible, and can match simple patterns. Clauses are separated at the beginning of lines that contain \verb/=>/ (where "lines" are also terminated by \verb/;/, and do not consider nested newlines), which avoids the need for any indentation-based parsing. For example:
\begin{lstlisting}
my-function = (
  ?x =>
    put one-argument %x
  ?y ?z =>
    put two-arguments
    put %y %z
  ... => ...
)
\end{lstlisting}

@@

\noindent
Functions maintain the scope of any free lexical variables in the body, including for mutation.

\subsubsection{Nestable Vectors}\noindent
We extend the concept of an \verb/argv/ vector such that it is nestable, using the compact syntax \verb/[ ]/. Combined with bareword strings (string literals without quotation), we can represent trees in a straightforward manner:
\begin{lstlisting}
a-tree = [node [node [leaf 1] [leaf 2]] [leaf 3]]
\end{lstlisting}
\noindent
These can be matched in lambda arguments by patterns such as \verb/[node ?x]/. A typical strategy for traversing this kind of structure might be a function that puts all leaf node values to its output in a predefined order, to be consumed by another process.

The builtin function \verb/for/ takes a vector as an argument and outputs each element in order, so that a vector can be traversed with:
\begin{lstlisting}
for [1 2 3 4 5] | each (?el => ...)
\end{lstlisting}

\subsubsection{Environments as Objects}\noindent
Variable environments can be captured directly as key-value maps with a parent pointer. Environment capture uses \verb/{ }/ syntax to run a block of code, then remove the running environment from its parent list. The resulting environment value is substituted in place of the \verb/{ }/ expression. Therefore all assignment syntaxes are available, including function definition. For example:
\begin{lstlisting}
(make-account ?balance) = {
 balance = $balance
 (deposit ?amt) = (%balance = (add %amt %balance))
 (withdraw ?amt) = (%balance = (sub %amt %balance))
}
\end{lstlisting}

\noindent
A planned extension would allow using special syntax to register additional parents, allowing users to inherit by direct delegation.

Environment lookup uses the \verb/!/ symbol, as in \verb/$env!key/. We use this symbol because it is already reserved by bash, unlike both \verb\/\ or \verb/./ which need to be available in bareword syntax to indicate file paths for external programs. Environments can also be mutated using the same access syntax. For example:
\begin{lstlisting}
account = (make-account 10)
$account!withdraw 4
put $account!balance # => 6
\end{lstlisting}

% \subsubsection{Reading and Writing}\noindent
% Reading and writing from the standard input and output are performed with the \verb/get/ and \verb/put/ functions, respectively. These block until they complete a communication.

\subsubsection{Collection and Substitution}\noindent
A priori, a capture mechanism such as described in Section \ref{return-semantics} would have to return a list, as any process may output zero or more values. However, as a function calling convention, that would require callers to manually unwrap lists on every function call.

In order to simplify substitution, Magritte uses normal parentheses to collect and \emph{expand} the resulting values into the current command vector---increasing the argument number by the number of values output from the function. For example:
\begin{lstlisting}
# A function defintion: output three values
(count-three) = (put 1; put 2; put 3)

# Collect three writes and expand 1 2 3 in-place
other-fn 0 (count-three) 4

@@

\noindent
These semantics are similar to the \verb/$(...)/ syntax in bash, with the exception that we have the ability to properly separate values without relying on whitespace.

@@

\section{Discussion}\label{discussion}
